### 2019-06-06 2-Chapters/2-Chapter/chapter2.tex
Done for a large proofreading tonight
Added leftbar to all defn,cor,thm etc

\paragraph{Abstract.}
%
In this chapter, we present the common base of all the models studied in this thesis: the multi-armed bandit (MAB) model, restricting to the single-player and stationary stochastic case. We focus on decision-making models with a finite number of resources, called arms, and on stochastic models, where an arm is associated with a one-dimensional distribution.
%
We formalize this MAB model, then we quote some of its important applications.
We define the notion of regret of an algorithm, and quote important results from the literature. A short review of the most important families of MAB algorithms is then given.
We conclude this chapter by discussing the question of online algorithm selection.

@@

Multi-Armed Bandits (MAB) models were introduced by Thompson as early as in 1933 \cite{Thompson33}, and later studied from the 1950s by Robbins \cite{Robbins52} and others.
This family of models was first proposed for clinical trials, and later applied to a wide range of different problems.
Their name refers to one-armed bandits found in casinos, as shown in Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
% The seminal paper by Lai and Robbins \cite{LaiRobbins85} was the first to introduce the measure of performance widely used in the literature, the notion of \emph{regret}, and to analyze a lower-bound on the performance of any algorithm against a fixed problem.
% Since then, many efficient algorithms were proposed, XXX
%
% In all this thesis, we only consider one-dimensional exponential families, and in particular Bernoulli distributions.
We formally define the model and the main notations used in all this document in Section~\ref{sec:2:notations}.
%
Multi-armed bandits have been applied to various fields, from recommender systems and information retrieval to healthcare and finance. We discuss some applications in Section~\ref{sec:2:applicationsofStochasticMAB}.
In the rest of this thesis, we focus on cognitive radio, an important field of research where bandits has already been applied with success in the last ten years.

@@

\paragraph{About notations.}
%
We use in this thesis the usual notations from the MAB literature,
% , and we follow closely the formalism of \cite{Kaufmann12PhD}, as well as \cite{Slivkins2019,LattimoreBanditAlgorithmsBook,Bubeck12}.
and a summary is included in the Nomenclature, in Page~\pageref{chapter:nomenclature}.

\paragraph{Main references for the curious reader.}
%
For more details and a more formal introduction to multi-armed bandits, we suggest to the interested reader to work on a very recent text-book by Slivkins \cite{Slivkins2019}.
Another excellent but reasonably short survey is the book by Bubeck and Cesa-Bianchi \cite{Bubeck12}, while the more recent book by Lattimore and Szepesv{\'a}ri \cite{LattimoreBanditAlgorithmsBook} is the most complete resource about bandit algorithms.
Finally, we recommend \cite{bouneffouf2019survey} for a short but good survey on applications of MAB.

@@

A MAB is a model of a decision-making game, where a \emph{player} has to sequentially select one action in a (usually finite) set of actions, and only receives a (random) feedback about the selected action, also called a \emph{reward}.
\textbf{We always consider $K\in\N, K\geq2$ arms, and discrete times $t\in\N^*$}.
When the player decides at time $t$ to play (or pull) the arm $A(t)\in\{1,\dots,K\}=[K]$, she receives a reward $r_{A(t)}(t)\in\R$, and so on for a finite number of steps $t$, from $t=1$ until $t=T$ for an \emph{horizon} $T$.
%
A commonly studied goal for the player is to maximize its sum of received rewards, $\sum_{t=1}^T r_{A(t)}(t)$
(or its \emph{mean} in stochastic and stationary models).

@@

\tikzstyle{block} = [align=center, draw, fill=gray!25, rectangle, minimum height=3em, minimum width=6em]
\begin{figure}[h!]
    \centering
    \resizebox{0.50\textwidth}{!}{
        \begin{tikzpicture}[auto, node distance=5cm, >=latex]
            %
            % We start by placing the blocks
            \node [block] (player) at (0,0) {Player};
            % We draw an edge between the player and system block to
            \node [block] (environment) at (4,0) {MAB problem};
            % Once the nodes are placed, connecting them is easy.
            \draw [->] (player) to[bend left=90] node[pos=0.5] {Chooses a \textbf{discrete} action $k=A(t) \in\{1,\dots,K\}$} (environment);
            \draw [->] (environment) to[bend left=90] node[pos=0.5] {Observes a \textbf{real-valued} reward $r_k(t) \in \R$} (player);
            %
        \end{tikzpicture}
    }
\caption{Reinforcement learning cycle in a MAB model, for time steps $t=1,\dots,T$.}
\label{fig:2:ReinforcementLearningCycleMABmodel}
\end{figure}

This quantity $\sum_{t=1}^T r_{A(t)}(t)$ can indeed be random, and it usually depends on two aspects.
First, the rewards may depend on the randomness of the environment, \ie, the unknown and unpredictable values $r_k(t)$.
Then this quantity also depends on the player's decision-making policy, \ie, the choices $A(t)$, that are based on (all) the \emph{past} observations, \ie, $A(1)$, $r_{A(1)}(1)$, $\dots$, $A(t-1)$, $r_{A(t-1)}(t-1)$, and possibly on an external source of randomness.
%
Without loss of generality, it can be given by a sequence $U_0,U_1,\dots$ of \iid{} random values, uniformly distributed on $[0,1]$, and that has to be independent from the rewards $r_k(t)$.
External randomness is used in most MAB algorithms, \eg, to break ties uniformly in an index policy.

@@

In all this thesis, we focus on the model with finite arms, and binary or real-valued stochastic rewards, that is $A(t)\in\{1,\dots,K\}$ for a fixed and known number of arms $K\in\N,K\geq2$, and $r_{A(t)}(t)\in\R$.
%
% Restricting to models with a finite number of arms
% rules out an entire domain of the recent research on MAB, in particular we do not consider linear nor contextual bandits (for more details, see Parts V and VI of \cite{LattimoreBanditAlgorithmsBook}).
%
We also restrict to \emph{stochastic} rewards, that is we associate a real-valued distribution to each arm, denoted $\nu_k$ for arm $k\in\{1,\dots,\}=[K]$.
Rewards are stationary, meaning that $(r_k(t))_{t\in\N^*}$ is independent and identically distributed (\iid), $r_k(t) \sim \nu_k$ for any $t\geq1$.

The most common objective for the player is to maximize its expected rewards $\E[\sum_{t=1}^T Y_t] = \sum_{t=1}^T \E[r_{A(t)}(t)]$.
The expectancy is taken on the rewards $(r_k(t))_{k,t}$, and the extern random variables $(U_s)_s$ (\ie, on the algorithm's decisions).
%
Other commonly studied objectives include best-arm identification (BAI) \cite{audibert2010best}.

@@

The webpage looks like Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
The arms follow Bernoulli distributions, \ie, $\nu_k = \mathcal{B}(\mu_k)$, of unknown means $\mu_k\in[0,1]$, and your goal in this interactive demonstration is to obtain the highest possible cumulated reward in $100$ steps, $\sum_{t=1}^{100} r_{A(t)}(t)$.
Your decisions have to be made sequentially: at time $t$, you pick one of the arms, $A(t) \in\{1,2,3,4,5\}$, then the demo shows the random reward obtained from this (virtual) casino machine (in \textcolor{gold}{yellow}), \ie, the binary value $r_{A(t)}(t)\in\{0,1\}$, that is sampled \iid{} from $\nu_k$.
%
The UI of the demo also shows the current value of $t$ (``total plays'') and $\sum_{s=1}^t r_{A(s)}(s)$ the ``total reward''.
For each arm, we show the sum of rewards obtained from that arm, \ie, $X_k(t) = \sum_{s=1}^t r_k(s) \mathbbm{1}(A(s) = k)$, in the ``Rewards'' line, and the number of pulls of that arm, \ie, $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ in the ``Pulls'' line.
%
The demo also shows the estimated probability of each arm, that is $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$), in the ``Estimated Probs'' line.

@@

A \emph{bandit algorithm} $\cA$ is also referred to as a strategy or a policy, and sometimes it is denoted by $\pi$ or $\rho$ in the literature.
The algorithm $\cA$ selects an arm $A(t)$ at time $t$, possibly by using the past observations and the past external randomness.
As shown below, being oblivious to the past yields very poor performance (cf. the pure exploration policy in Section~\ref{sub:2:naiveSimpleStrategies}), so efficient policies indeed depend on the successive feedbacks.
%
More formally, an algorithm can be defined as a sequence of \emph{measurable} functions $(\cA_t)_{t\geq1}$,
where $\cA_t$ maps the past observations $O_t = (U_0, Y_1, U_1, \dots, Y_{t-1}, U_{t-1})$
to an arm $\cA_t(O_t) = A(t) \in[K]$,
if we denote $Y_s = r_{A(s)}(s)$ the $s$-th reward.
The initial information is reduced to $O_1 = (U_0)$, and the first decision is $A(1) = \cA_1(O_1)$. Usually, most algorithms starts by selecting $A(1)=1,\dots,A(K)=K$ (or a permutation of the $K$ arms) in the $K$ first steps (\eg, Algorithm~\ref{algo:2:indexPolicy}).
% as illustrated for example in Algorithm~\ref{algo:2:indexPolicy} below.
%
An algorithm is said to be \emph{deterministic} if it does not depend on the external randomness $U_0,U_1,\dots$, but in this thesis \textbf{we only use non-deterministic algorithms}
(in particular, index policies need $U_t$ to break ties in the $\argmax$, see Algorithm~\ref{algo:2:indexPolicy} below).

@@

In the example above, we consider Bernoulli distributions, but other real-valued distributions have been studied in the literature.
From now on and until the last chapter of this thesis, \textbf{we only consider stochastic rewards}. The piece-wise stationary model is studied in Chapter~\ref{chapter:6}.
%
% Everything is implemented and documented at
% https://smpybandits.github.io/docs/Arms.html
\textbf{We also focus only on real-valued rewards}, meaning that $r_k(t)\in\R$ for all arm $k$ and time $t$.

@@

\begin{defn}[Kullback-Leibler divergence]\label{def:2:KLDivergence}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    The KL divergence between two distributions $d_1$ and $d_2$ is defined as
    \begin{align}\label{eq:2:Kullback-LeiblerDivergenceExpFamily1}
        \KL\left(d_1, d_2\right) = \int d_1 \log\left(\frac{d_2}{d_1}\right) \lambda(\mathrm{d}x) = \E_{d_1} \left[ \log\left(\frac{d_2}{d_1}\right) \right] \in \mathbb{R}\cup\{\pm\infty\}.
    \end{align}
\vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{defn}

@@

Previous works of our SCEE team showed that MAB can be used to model the problem of spectrum access for a secondary user accessing a licensed spectrum.
In this model, arms represent a finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
In the model with sensing, the rewards $r_k(t)$ represents the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$.
A reward of $r_k(t) = 1$ indicates that no Primary User was sensed, while a reward of $r_k(t)=0$ indicates that the channel $k$ is busy at time $t$ and no uplink message can be sent.
%
This model was first introduced by Wassim Jouini during his PhD thesis,
in \cite{Jouini09} and later studied in both \cite{Jouini10,Jouini12}.
Proof-of-concepts using real-world radio hardware was first propsed in \cite{MoyWSR2014,RobertSDR2014}.

@@

Let us first introduce some notations.
From now on, we consider a stochastic and stationary MAB problem, with $K$ arms of distributions $\nu_1,\dots,\nu_K=(\nu_k)_k$, that generates \iid{} rewards $r_k(t)\sim \nu_k$, for any time $t$.
We focus on distributions fully characterized by their means, that can be Bernoulli or any one-dimensional exponential family.
We denote $\mu_k = \E[r_k(t)] \in \R$ the mean of the distribution of arm $k$ (it will be referred to as the \emph{mean of arm $k$}).
%
% \paragraph{Defining the regret.}
%
To define the regret, we first need to distinguish between \emph{optimal} and \emph{sub-optimal} arms.

\begin{defn}\label{def:2:optimalSubOptimalArms}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    % [Optimal and sub-optimal arms]
    Consider a bandit problem of $K$ arms with distributions of means $\bm{\mu}=\mu_1,\dots,\mu_k$.
    Denote $\mu^* = \max_k \mu_k$ the largest mean.
    The best arm can be non unique, and any arm $k$ having $\mu_k = \mu^*$ is said to be \emph{optimal},
    while arms satisfying $\mu_k < \mu^*$ are called \emph{sub-optimal}.
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{defn}

If the goal of the player is to maximize $\E\left[\sum_{t=1}^T r_{A(t)}(t)\right]$,
the optimal strategy for this bandit problem is to always pull one of the optimal arms (it can be not unique), but it is unrealistic as the player does not know the true means, nor the optimal arms (as long as they are not all optimal, \ie, in non trivial problems).
%
Comparing the difference between the performance of a fixed baseline and that of the player is a common approach in machine learning research,
and here we can compare with the oracle strategy that always obtains an (expected) reward of $\mu^*$.
%
For a fixed horizon $T$, if $k^*$ denotes the index of any optimal arm,
let us introduce the (mean) regret $R_T^{\cA}$ of an algorithm $\cA$ as
$R_T^{\cA} = \E\left[ \sum_{t=1}^T (r_{k^*}(t) - r_{A(t)}(t)) \right]$.
%
As the rewards from arm $k^*$ are \iid{} (as for all arms), and by linearity of the expectancy, we can rewrite this expression to obtain the following definition of the regret:
\begin{align*}
    R_T^{\cA}
    & = \E\left[ \sum_{t=1}^T (r_{k^*}(t) - r_{A(t)}(t)) \right]
    = \E\left[ \sum_{t=1}^T \underbrace{r_{k^*}(t)}_{\text{\iid{} from $\nu_{k^*}$}} \right] - \E\left[ \sum_{t=1}^T r_{A(t)}(t) \right] \\
    & = T \times \E\left[ r_{k^*}(t=1) \right] - \sum_{t=1}^T \E\left[ r_{A(t)}(t) \right]
    = T \mu^* - \sum_{t=1}^T \E\left[ r_{A(t)}(t) \right].
\end{align*}
% This is the definition we give below.

\begin{defn}[Regret]\label{def:2:regret}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    For an algorithm $\cA$, a bandit problem of $K$ arms characterized by their means $\bm{\mu} = \mu_1,\dots,\mu_K$, with $\mu^* = \max_k \mu_k$, then its \emph{(mean) regret} at horizon $T$ is $R_T^{\cA}$ defined as
    %
    \begin{align}\label{eq:2:regret}
        R_T^{\cA}
        = \E\left[ \sum_{t=1}^T (r_{k^*}(t) - r_{A(t)}(t)) \right]
        % = T \mu^* - \sum_{t=1}^T \E\left[ Y(t) \right]
        = T \mu^* - \sum_{t=1}^T \E\left[ r_{A(t)}(t) \right].
    \end{align}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{defn}

\paragraph{A useful decomposition of the regret.}
%
Remember that $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ denotes the number of times arm $k$ was selected between times $1$ and $t$,
%
and that the rewards $r_k(t)$ are all \iid{} of mean $\mu_k$.
The \emph{gap} between any arm $k\in[K]$ and an optimal arm is defined as $\Delta_k = \mu^* - \mu_k$
(an arm $k$ is thus sub-optimal if and only if $\Delta_k > 0$),
and thus we can write the following decomposition on the regret.

>>>>

\textbf{Abstract.}
%
In this chapter, we present the common base of all the models studied in this thesis: the multi-armed bandit (MAB) model, restricting to the single-player and stationary stochastic case. We focus on decision-making models with a finite number of resources, called arms, and on stochastic models, where an arm is associated with a one-dimensional distribution.
%
We formalize this MAB model, then we quote some of its important applications.
We define the notion of regret of an algorithm, and quote important results from the literature. A short review of the most important families of MAB algorithms is then given.
We conclude this chapter by discussing the question of online algorithm selection.

@@

Multi-Armed Bandits (MAB) models were introduced by Thompson as early as in 1933 \cite{Thompson33}, and later studied from the 1950s by Robbins \cite{Robbins52} and others.
This family of models was first proposed for clinical trials, and later applied to a wide range of different problems.
Their name refers to one-armed bandits found in casinos, as shown in Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
% The seminal paper by Lai and Robbins \cite{LaiRobbins85} was the first to introduce the measure of performance widely used in the literature, the notion of \emph{regret}, and to analyze a lower-bound on the performance of any algorithm against a fixed problem.
% Since then, many efficient algorithms were proposed, XXX

% In all this thesis, we only consider one-dimensional exponential families, and in particular Bernoulli distributions.
We formally define the model and the main notations used in all this document in Section~\ref{sec:2:notations}.
%
Multi-armed bandits have been applied to various fields, from recommender systems and information retrieval to healthcare and finance. We discuss some applications in Section~\ref{sec:2:applicationsofStochasticMAB}.
In the rest of this thesis, we focus on cognitive radio, an important field of research where bandits has already been applied with success in the last ten years.

@@

\textbf{About notations.}
%
We use in this thesis the usual notations from the MAB literature,
% , and we follow closely the formalism of \cite{Kaufmann12PhD}, as well as \cite{Slivkins2019,LattimoreBanditAlgorithmsBook,Bubeck12}.
and a summary is included in the Nomenclature, in Page~\pageref{chapter:nomenclature}.

\textbf{Main references for the curious reader.}
%
For more details and a more formal introduction to multi-armed bandits, we suggest to the interested reader to work on a very recent text-book by Slivkins \cite{Slivkins2019}.
Another excellent but reasonably short survey is the book by Bubeck and Cesa-Bianchi \cite{Bubeck12}, while the more recent book by Lattimore and Szepesv{\'a}ri \cite{LattimoreBanditAlgorithmsBook} is the most complete resource about bandit algorithms.
Finally, we recommend \cite{bouneffouf2019survey} for a short but good survey on applications of MAB.

@@

A MAB is a model of a decision-making game, where a \emph{player} has to sequentially select one action in a (usually finite) set of actions, and only receives a (random) feedback about the selected action, also called a \emph{reward}.
\textbf{We always consider $K\in\N, K\geq2$ arms, and discrete times $t\in\N^*$}.
When the player decides at time $t$ to play (or pull) the arm $A(t)\in\{1,\dots,K\}=[K]$, she receives a reward $r(t) = Y_{A(t),t}\in\R$, and so on for a finite number of steps $t$, from $t=1$ until $t=T$ for an \emph{horizon} $T$.
%
A commonly studied goal for the player is to maximize its sum of received rewards, $\sum_{t=1}^T r(t)$
(or its \emph{mean} in stochastic and stationary models).

@@

\tikzstyle{block} = [align=center, draw, fill=gray!25, rectangle, minimum height=3em, minimum width=6em]
\begin{figure}[h!]
    \centering
    \resizebox{0.50\textwidth}{!}{
        \begin{tikzpicture}[auto, node distance=5cm, >=latex]
            %
            % We start by placing the blocks
            \node [block] (player) at (0,0) {Player};
            % We draw an edge between the player and system block to
            \node [block] (environment) at (4,0) {MAB problem};
            % Once the nodes are placed, connecting them is easy.
            \draw [->] (player) to[bend left=90] node[pos=0.5] {Chooses a \textbf{discrete} action $k=A(t) \in\{1,\dots,K\}$} (environment);
            \draw [->] (environment) to[bend left=90] node[pos=0.5] {Observes a \textbf{real-valued} reward $r(t)=Y_{k,t} \in \R$} (player);
            %
        \end{tikzpicture}
    }
\caption{Reinforcement learning cycle in a MAB model, for time steps $t=1,\dots,T$.}
\label{fig:2:ReinforcementLearningCycleMABmodel}
\end{figure}

This quantity $\sum_{t=1}^T r(t)$ can indeed be random, and it usually depends on two aspects.
First, the rewards may depend on the randomness of the environment, \ie, the unknown and unpredictable values $r(t)=Y_{k,t}$.
Then this quantity also depends on the player's decision-making policy, \ie, the choices $A(t)$, that are based on (all) the \emph{past} observations, \ie, $A(1)$, $Y_{A(1),1}$, $\dots$, $A(t-1)$, $Y_{A(t-1),t-1}$, and possibly on an external source of randomness.
%
Without loss of generality, it can be given by a sequence $U_0,U_1,\dots$ of \iid{} random values, uniformly distributed on $[0,1]$, and that has to be independent from the samples $Y_{k,t}$.
External randomness is used in most MAB algorithms, \eg, to break ties uniformly in an index policy.

@@

In all this thesis, we focus on the model with finite arms, and binary or real-valued stochastic rewards, that is $A(t)\in\{1,\dots,K\}$ for a fixed and known number of arms $K\in\N,K\geq2$, and $Y_{A(t),t}\in\R$.
%
% Restricting to models with a finite number of arms
% rules out an entire domain of the recent research on MAB, in particular we do not consider linear nor contextual bandits (for more details, see Parts V and VI of \cite{LattimoreBanditAlgorithmsBook}).
%
We also restrict to \emph{stochastic} rewards, that is we associate a real-valued distribution to each arm, denoted $\nu_k$ for arm $k\in\{1,\dots,\}=[K]$.
Rewards are stationary, meaning that $(Y_{k,t})_{t\in\N^*}$ is independent and identically distributed (\iid), $Y_{k,t} \sim \nu_k$ for any $t\geq1$.

The most common objective for the player is to maximize its expected rewards $\E[\sum_{t=1}^T r(t)] = \sum_{t=1}^T \E[Y_{A(t),t}]$.
The expectancy is taken on the rewards $(Y_{k,t})_{k,t}$, and the extern random variables $(U_s)_s$ (\ie, on the algorithm's decisions).
%
Other commonly studied objectives include best-arm identification (BAI) \cite{audibert2010best}.

% \TODOL{Je dois unifier les notations, et utiliser $Y_{k,t}$ pour les flux qui sont iid et $r(t)$ pour les récompenses effectivement observées (comme dans les chapitres 5 et 6)}

@@

The webpage looks like Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
The arms follow Bernoulli distributions, \ie, $\nu_k = \mathcal{B}(\mu_k)$, of unknown means $\mu_k\in[0,1]$, and your goal in this interactive demonstration is to obtain the highest possible cumulated reward in $100$ steps, $\sum_{t=1}^{100} r(t)$.
Your decisions have to be made sequentially: at time $t$, you pick one of the arms, $A(t) \in\{1,2,3,4,5\}$, then the demo shows the random reward obtained from this (virtual) casino machine (in \textcolor{gold}{yellow}), \ie, the binary value $r(t)\in\{0,1\}$, that is sampled \iid{} from $\nu_k$.
%
The UI of the demo also shows the current value of $t$ (``total plays'') and $\sum_{s=1}^t r(s)$ the ``total reward''.
For each arm, we show the sum of rewards obtained from that arm, \ie, $X_k(t) = \sum_{s=1}^t r(s) \mathbbm{1}(A(s) = k)$, in the ``Rewards'' line, and the number of pulls of that arm, \ie, $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ in the ``Pulls'' line.
%
The demo also shows the estimated probability of each arm, that is $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$), in the ``Estimated Probs'' line.

@@

A \emph{bandit algorithm} $\cA$ is also referred to as a strategy or a policy, and sometimes it is denoted by $\pi$ or $\rho$ in the literature.
The algorithm $\cA$ selects an arm $A(t)$ at time $t$, possibly by using the past observations and the past external randomness.
As shown below, being oblivious to the past yields very poor performance (cf. the pure exploration policy in Section~\ref{sub:2:naiveSimpleStrategies}), so efficient policies indeed depend on the successive feedbacks.
%
More formally, an algorithm can be defined as a sequence of \emph{measurable} functions $(\cA_t)_{t\geq1}$,
where $\cA_t$ maps the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$
to an arm $\cA_t(O_t) = A(t) \in[K]$
(we remind that we denote $r(s) = Y_{A(s),s}$ the $s$-th reward).
The initial information is reduced to $O_1 = (U_0)$, and the first decision is $A(1) = \cA_1(O_1)$. Usually, most algorithms starts by selecting $A(1)=1,\dots,A(K)=K$ (or a permutation of the $K$ arms) in the $K$ first steps (\eg, Algorithm~\ref{algo:2:indexPolicy}).
% as illustrated for example in Algorithm~\ref{algo:2:indexPolicy} below.
%
An algorithm is said to be \emph{deterministic} if it does not depend on the external randomness $U_0,U_1,\dots$, but in this thesis \textbf{we only use non-deterministic algorithms}
(in particular, index policies need $U_t$ to break ties in the $\argmax$, see Algorithm~\ref{algo:2:indexPolicy} below).

@@

In the example above, we consider Bernoulli distributions, but other real-valued distributions have been studied in the literature.
From now on and until the last chapter of this thesis, \textbf{we only consider stochastic rewards}. The piece-wise stationary model is studied in Chapter~\ref{chapter:6}.
%
% Everything is implemented and documented at
% https://smpybandits.github.io/docs/Arms.html
\textbf{We also focus only on real-valued rewards}, meaning that $Y_{k,t}\in\R$ for all arm $k$ and time $t$.

@@

\begin{definition}[Kullback-Leibler divergence]\label{def:2:KLDivergence}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    The KL divergence between two distributions $d_1$ and $d_2$ is defined as
    \begin{align}\label{eq:2:Kullback-LeiblerDivergenceExpFamily1}
        \KL\left(d_1, d_2\right) = \int d_1 \log\left(\frac{d_2}{d_1}\right) \lambda(\mathrm{d}x) = \E_{d_1} \left[ \log\left(\frac{d_2}{d_1}\right) \right] \in \mathbb{R}\cup\{\pm\infty\}.
    \end{align}
\vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

@@

Previous works of our SCEE team showed that MAB can be used to model the problem of spectrum access for a secondary user accessing a licensed spectrum.
In this model, arms represent a finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
In the model with sensing, the samples $Y_{k,t}$ represents the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$.
A reward of $r(t) = 1$ indicates that no Primary User was sensed, while a reward of $r(t)=0$ indicates that the channel $k$ is busy at time $t$ and no uplink message can be sent.
%
This model was first introduced by Wassim Jouini during his PhD thesis,
in \cite{Jouini09} and later studied in both \cite{Jouini10,Jouini12}.
Proof-of-concepts using real-world radio hardware was first propsed in \cite{MoyWSR2014,RobertSDR2014}.

@@

Let us first introduce some notations.
From now on, we consider a stochastic and stationary MAB problem, with $K$ arms of distributions $\nu_1,\dots,\nu_K=(\nu_k)_k$, that generates \iid{} samples $Y_{k,t} \sim \nu_k$, for any time $t$.
% with $r(t)=Y_{A(t),t}$).
We focus on distributions fully characterized by their means, that can be Bernoulli or any one-dimensional exponential family.
We denote $\mu_k = \E[Y_{k,t}] \in \R$ the mean of the distribution of arm $k$ (it will be referred to as the \emph{mean of arm $k$}).
%
% \paragraph{Defining the regret.}
%
To define the regret, we first need to distinguish between \emph{optimal} and \emph{sub-optimal} arms.

\begin{definition}\label{def:2:optimalSubOptimalArms}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    % [Optimal and sub-optimal arms]
    Consider a bandit problem of $K$ arms with distributions of means $\bm{\mu}=\mu_1,\dots,\mu_k$.
    Denote $\mu^* = \max_k \mu_k$ the largest mean.
    The best arm can be non unique, and any arm $k$ having $\mu_k = \mu^*$ is said to be \emph{optimal},
    while arms satisfying $\mu_k < \mu^*$ are called \emph{sub-optimal}.
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

If the goal of the player is to maximize $\E\left[\sum_{t=1}^T r(t)\right]$,
the optimal strategy for this bandit problem is to always pull one of the optimal arms (it can be not unique), but it is unrealistic as the player does not know the true means, nor the optimal arms (as long as they are not all optimal, \ie, in non trivial problems).
%
Comparing the difference between the performance of a fixed baseline and that of the player is a common approach in machine learning research,
and here we can compare with the oracle strategy that always obtains an (expected) reward of $\mu^*$.
%
For a fixed horizon $T$, if $k^*$ denotes the index of any optimal arm,
let us introduce the (mean) regret $R_T^{\cA}$ of an algorithm $\cA$ as
$R_T^{\cA} = \E\left[ \sum_{t=1}^T (Y_{k^*,t} - r(t)) \right]$.
%
As the rewards from arm $k^*$ are \iid{} (as for all arms), and by linearity of the expectancy, we can rewrite this expression to obtain the following definition of the regret:
\begin{align*}
    R_T^{\cA}
    & = \E\left[ \sum_{t=1}^T (Y_{k^*,t} - r(t)) \right]
    = \E\left[ \sum_{t=1}^T \underbrace{Y_{k^*,t}}_{\text{\iid{} from $\nu_{k^*}$}} \right] - \E\left[ \sum_{t=1}^T r(t) \right] \\
    & = T \times \E\left[ Y_{k^*,1} \right] - \sum_{t=1}^T \E\left[ r(t) \right]
    = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right].
\end{align*}
% This is the definition we give below.

\begin{definition}[Regret]\label{def:2:regret}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    For an algorithm $\cA$, a bandit problem of $K$ arms characterized by their means $\bm{\mu} = \mu_1,\dots,\mu_K$, with $\mu^* = \max_k \mu_k$, then its \emph{(mean) regret} at horizon $T$ is $R_T^{\cA}$ defined as
    %
    \begin{align}\label{eq:2:regret}
        R_T^{\cA}
        = \E\left[ \sum_{t=1}^T (Y_{k^*,t} - r(t)) \right]
        % = T \mu^* - \sum_{t=1}^T \E\left[ Y(t) \right]
        = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right].
    \end{align}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

\paragraph{A useful decomposition of the regret.}
%
Remember that $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ denotes the number of times arm $k$ was selected between times $1$ and $t$,
%
and that the samples $Y_{k,t}$ are all \iid{} of mean $\mu_k$.
The \emph{gap} between any arm $k\in[K]$ and an optimal arm is defined as $\Delta_k = \mu^* - \mu_k$
(an arm $k$ is thus sub-optimal if and only if $\Delta_k > 0$),
and thus we can write the following decomposition on the regret.

\begin{lemma}[Regret decomposition]\label{lem:2:RegretDecomposition}
\begin{leftbar}[lemmabar]  % XXX leftbar lemmabar, comment if needed
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$, weighted by their gaps:
    \begin{align}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{align}
    \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar lemmabar, comment if needed
\end{lemma}
%
\begin{proof}\label{proof:2:RegretDecomposition}
    We use the chain rule of expectation and a conditioning on $O_t$,
    because the expectation is taken on the randomness of the (\iid) samples $(Y_{k,t})_t$ and on the decisions of the player $(A(t))_t$, which are measurable wrt to the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$.
    Thus we can rewrite the expected cumulated reward as follows:
    %
    % FIXME add a blank line or a negative space if it looks weird
    %
    % \vspace*{-20pt}
    \begin{align*}
        \E \left[ \sum_{t=1}^T r(t) \right]
        &= \E \left[ \sum_{t=1}^T \sum_{k=1}^K Y_{k,t} \mathbbm{1}(A(t) = k) \right]
        = \sum_{k=1}^K \sum_{t=1}^T \E \left[ Y_{k,t} \mathbbm{1}(A(t) = k) \right] \\
        &= \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,t} | O_t \right] \mathbbm{1}(A(t) = k) \Bigr]
        = \sum_{k=1}^K \sum_{t=1}^T \E \Bigl[ \E \left[ Y_{k,1} \right] \mathbbm{1}(A(t) = k) \Bigr] \\
        &= \sum_{k=1}^K \underbrace{\E \left[ Y_{k,1} \right]}_{\mu_k} \underbrace{\sum_{t=1}^T \E \left[ \mathbbm{1}(A(t) = k) \right]}_{= N_k(T)}
        % = \sum_{k=1}^K \mu_k \E \left[ \sum_{t=1}^T \mathbbm{1}(A(t) = k) \right]\\
        = \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].
    \end{align*}
    %
    Thus $R_T^{\cA} = T \mu^* - \sum_{t=1}^T \E\left[ r(t) \right] = T \mu^* - \sum_{k=1}^K \mu_k \E[ N_k(T) ] = \sum_{k=1}^K (\mu^* - \mu_k) \E[ N_k(T) ]$,
    as $\sum_{k=1}^K  \E[ N_k(T) ] = T$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{proof}
